---
title: "Class08: Principal Component Analysis (PCA)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PCA of UK food data

```{r}
x <- read.csv("UK_foods.csv")
```

Q. How many rows and columns are in "UK_foods.csv"?
```{r}
dim(x)
```

```{r}
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```


```{r}
pca <- prcomp(t(x))
summary(pca)
```

# Generate a plot of PC1 vs PC2. 
```{r}
plot(pca$x[,1],pca$x[,2],xlab="PC1",ylab="PC2",xlim=c(-270,500))
text(pca$x[,1],pca$x[,2], colnames(x), col=c("orange","red","blue","darkgreen"))
```

# Examine the "loadings"

This will help to determine how the origional variable (dimensions) contribute to the new PCs

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```

```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```

## PCA of RNA-seq data

```{r}
rna.data <- read.csv("expression.csv", row.names=1)
head(rna.data)
```

# Generating plots for RNA data

```{r}
pca <- prcomp(t(rna.data), scale=TRUE)
 
plot(pca$x[,1], pca$x[,2])

#This plot is made to separate the samples into two groups of 5 samples each. 
```

```{r}
# Calculate how much variation in the original data
pca.var <- pca$sdev^2

pca.var.per <- round(pca.var/sum(pca.var)*100,1)
pca.var.per
```

# Generate scree-plot 
```{r}
barplot(pca.var.per, main="Scree Plot", xlab="Principal Component", ylab="Percent Variation")

#This scree-plot shows that PC1 is where all the action is.  
```

# Making PCA plot look more efficient
```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <-"blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16, xlab=paste0("PC1 (",pca.var.per[1], "%)"),ylab=paste0("PC2 (",pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

# Find the top 10 genes that contribute most to PC1 in either positive or negative way. 

```{r}
loading_scores <- pca$rotation[,1]

gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

```

# The names of the top 10 genes are listed below. 
```{r}
top_10_genes <- names(gene_score_ranked[1:10])

top_10_genes 

```